---
title: "Predicting effect of CRISPR KOs by GRP decomposition"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced usage: Predicting effect of CRISPR KOs (Figure 3G)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=F, message=F}
require(Seurat)
require(hsc19)
require(pheatmap)
require(glmnet)
require(reshape2)
require(plyr)
require(ggrepel)
set.seed(1337)
```

## Summary

This vignette decomposes transcriptomes of Lin-Sca1+Kit+ and Lin-Sca1-Kit cells into 19 GRPs, and uses this decomposition to identify regulators of stemness exit (i.e. Sca1-positivity). It then compares these predictions to a [published CRISPR screen for Sca1 expression](https://www.nature.com/articles/s41588-023-01471-2)

## Data prepatation

We download a scRNA-seq dataset of Lin-c-Kit+ cells (from [Scherer et al., 2025](https://www.nature.com/articles/s41586-025-09041-8)), and use CITE-seq data to annotate cells as Sca1+ or Sca1-

```{r prepareSeurat, eval=T, warning=F, message=F}

if (!file.exists("data_LARRY/seurat_LK_LSK.RDS")) download.file(
  "https://ndownloader.figshare.com/files/42587815",
  destfile = "data_LARRY/seurat_LK_LSK.RDS",
  mode = "wb"
)

larry <- readRDS("data_LARRY/seurat_LK_LSK.RDS")
larry <- subset(larry, nFeature_AB > 15)
#FeatureScatter(larry, "SCA1-totalseqb", "cKIT-totalseqb",split.by =  "Experiment")
ABs <- GetAssayData(larry, assay = "AB")
larry$gate <- ifelse(larry$Experiment == "LARRY replicate",
                     ifelse(ABs["cKIT-totalseqb",] < 3, "Kit-", ifelse(ABs["SCA1-totalseqb",] > 3, "LSK", "LK")),
                     ifelse(ABs["cKIT-totalseqb",] < 2, "Kit-", ifelse(ABs["SCA1-totalseqb",] > 2, "LSK", "LK")))
#DimPlot(larry, group.by = "gate")
#DimPlot(larry)
larry$randomid <- rep(1:5, length.out = ncol(larry))

#make pseudobulks for LK and LSK. Then, decompose those to learn a model for "Sca positivity"
pseudobulks <- AggregateExpression(larry, return.seurat = T, group.by = c("gate", "ProcessingBatch", "randomid"))
VariableFeatures(pseudobulks) <- VariableFeatures(larry)
```

## Decompose data into GRPs

```{r decompose}
decomp <- decompose(s = pseudobulks)
col.labels <- data.frame(row.names = colnames(decomp),
                         gate = gsub("_.+", "", colnames(decomp)),
                         mouse = gsub(".+_(.+)_.+", "\\1", colnames(decomp)))
```

## Build model (Figure S4C,D)

How do GRPs affect the phenotype (i.e. Sca1 positivity)? Build a linear model and check on a held-out mouse (that mouse was processed on a different day from the remaining mice) - Figure S4D

```{r buildmodel, fig.width=2.8, fig.height=2.5 }
#now, build model
features <- decomp@result[,col.labels$gate != "Kit-" & col.labels$mouse != "LK-LSK-stained"]
response <- grepl("^LSK", colnames(features))
linear <- glm(response ~ ., family = "binomial", data = data.frame(response=response, t(features[-1,])))

new.features <- decomp@result[,col.labels$gate != "Kit-" & col.labels$mouse == "LK-LSK-stained"]
new.response <-  grepl("^LSK", colnames(new.features))

for_model_plot <- data.frame(cellstate = c(response,new.response), predicted = c(predict(linear, type = "response"), predict(linear, as.data.frame(t(new.features)),type = "response")),
                             set = c(rep("Mouse1-4\n(train data)", length(response)), rep("Mouse5\n(test data)", length(new.response))))

figs4d <- ggplot(aes(x = cellstate, y = predicted), data =for_model_plot) + facet_wrap(~set) + geom_point(position = position_jitter(width=0.4)) +
  theme_bw(base_size = 14) + theme(panel.grid = element_blank()) + scale_x_discrete(name = "Cell type", labels = c("FALSE" = "LK", "TRUE" = "LSK")) + ylab("Estimated probability LSK")
print(figs4d)
```

Extract and vidualize model coefficients - Figure S4C:

```{r modelcoef, fig.width=1.1, fig.height=2.8}
linear_coefs <- coef(linear)[as.character(factor_annotation$factor_id)]
linear_coefs_forplot <- matrix(linear_coefs, ncol=1, dimnames = list(gsub("Factor_","",names(linear_coefs)), NULL))
pheatmap(linear_coefs_forplot, border_color =NA, cluster_cols = F, show_colnames = F, cluster_rows = F)
```

## Predict CRISPR screen

Here we multiply the effect of the gene KOs on the GRPs (beta matrix from GSFA, summarised by gene) with the effect of the GRP on the phenotype, and we compare these predictions to the published CRISPR KO screen result [David Lara et al, Table S3](https://www.nature.com/articles/s41588-023-01471-2)

```{r predict, fig.width=2.8, fig.height=2.5}

david <- read.csv("data_Lara_et_al/david_tableS3.csv", header=T)[,1:6]
lsk_screen <- subset(david, Population.2 == "LSKd9" & Population.1 == "cKit")
our_genes_subset <- unique(gsub("-.+", "", rownames(hsc18_perturbation_beta)))

#make a matrix: Mean of significant guides for each gene
beta_sparse_long <- melt(beta_mat_sparse, varnames = c("guide_id", "factor_id"), value.name = "sparse")
beta_sparse_long$guide_id <- gsub("-CRISPRi-sgRNA-", ".", beta_sparse_long$guide_id)

beta_full_long <- melt(beta_mat, varnames = c("guide_id", "factor_id"), value.name = "full")
beta_full_long$guide_id <- gsub("-CRISPRi-sgRNA-", ".", beta_full_long$guide_id)

gamma_long <- melt(gamma_mat, varnames = c("guide_id", "factor_id"), value.name = "gamma")
gamma_long$guide_id <- gsub("-CRISPRi-sgRNA-", ".", gamma_long$guide_id)

factor_regulators <- merge(factor_annotation, beta_sparse_long)
factor_regulators <- merge(factor_regulators, beta_full_long)
factor_regulators <- merge(factor_regulators, gamma_long)
factor_regulators$gene <- gsub("\\.\\d$", "", factor_regulators$guide_id)
factor_genes <- ddply(factor_regulators, c("gene", "factor_id", "factor_label", "multiply_with"), summarise, 
                      n.sig = sum(sparse != 0), 
                      rel.sig = sum(sparse != 0) / length(sparse), 
                      mean.sig = mean(sparse[sparse!=0]), mean.full = mean(full))
factor_genes$mean.sig[is.na(factor_genes$mean.sig)] <- 0
predictor_matrix <- dcast(factor_genes, gene ~ factor_id, value.var = "mean.full")
predictor_matrix[is.na(predictor_matrix)] <- 0
rownames(predictor_matrix) <- predictor_matrix$gene
predictor_matrix <- predictor_matrix[,names(linear_coefs)]
overlapping_genes <- intersect(our_genes_subset, lsk_screen$Gene)
#for each overlapping gene
predicted <- sapply(overlapping_genes, function(g) {
  #compute effect on factors and multiply with coefs
  sum(predictor_matrix[g,] * linear_coefs)
})
lsk_screen$predicted <- predicted[as.character(lsk_screen$Gene)]

forplot <- subset(lsk_screen, Significant.guides >0 & !is.na(predicted))
forplot$discrete <- infotheo::discretize(forplot$predicted,nbins=3)$X
ctest <- with(forplot, cor.test(predicted, Normalized.lineage.score))

fig3g <- ggplot(aes(x = predicted, y = Normalized.lineage.score), data= forplot) + theme_bw(base_size = 14) + theme(panel.grid = element_blank()) + geom_point() + geom_hline(yintercept = -2, linetype=3) + geom_vline(xintercept = 0, linetype=3) +
  xlab("Predicted effect on\nLSK/LK ratio") + ylab("CRISPR\nLSK/LK score") + geom_text(aes(label = l, x=x, y=y), data =data.frame(l=sprintf("R=%.2f\np=%.4f", ctest$estimate, ctest$p.value), x = -35, y = 2.5)) +
  geom_text_repel(aes(label = Gene), subset(forplot, Gene %in% c("Max", "Ash2l","Rbbp5")))

print(fig3g)
```



