---
title: "Decomposition of human ageing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Basic package usage: Human Ageing (Figure 3B,C)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=F, message=F}
library(hsc19)
library(Seurat)
library(ggplot2)
```

## Summary

This vignette decomposes HSC transcriptomes from [a large scale resource on human ageing](https://www.nature.com/articles/s41591-025-03716-5) into nineteen gene regulatory programs. It demonstrates the basic use of the `decompose` and `associate` functions, and reproduces figures 3B and C from the manuscript.

## Data preparation

Step 1 (not run): Download data from [CellXGene](https://cellxgene.cziscience.com/collections/5542eeb0-96ef-4ab9-95ea-eb6abc178461)

```{bash download, eval=F}
wget -O furer_ultima.h5ad https://datasets.cellxgene.cziscience.com/fd9e6b84-4687-44cb-a61e-1b261c05c414.h5ad
wget -O furer_illumina.h5ad https://datasets.cellxgene.cziscience.com/00414a7b-5aba-4c51-ad1e-d6caa6b1ebf3.h5ad
```

Step 2 (not run): Create patient-level pseudobulks of HSCs - shown for ultima data

```{python pseudobulk, eval=F}

import scanpy as sc
import numpy as np
from scipy import sparse
import pandas as pd
ad = sc.read("furer_ultima.h5ad")
hsc_normal = ad[(ad.obs["cell state"] == "HSC_MPP") & (ad.obs["disease"] == "normal")]


# Extract donor IDs
donor_ids = ad.obs["donor_id"].values

# Get unique donors
unique_donors = np.unique(donor_ids)

# Create pseudobulks
pseudobulk_list = []

for donor in unique_donors:
    print(donor + "\n")
    # Select cells for this donor
    donor_mask = donor_ids == donor
    donor_counts = ad[donor_mask].X
    # Sum counts across cells
    if sparse.issparse(donor_counts):
        summed_counts = np.array(donor_counts.sum(axis=0)).flatten()
    else:
        summed_counts = donor_counts.sum(axis=0)
    pseudobulk_list.append(summed_counts)
    
pseudobulk_matrix = pd.DataFrame(
    np.vstack(pseudobulk_list),
    index=unique_donors,
    columns=ad.var_names
)

pseudobulk_matrix.transpose().to_csv("Furer_et_al_ultima_pseudobulks.csv.gz", compression='gzip')
```

Step 3: Load pseudobulks in R, create Seurat object

```{r prepareSeurat, eval=T}
illumina_pseudobulks <- read.csv("data_ageing/Furer_et_al_illumina_pseudobulks.csv.gz", row.names = 1)
ultima_pseudobulks <- read.csv("data_ageing/Furer_et_al_ultima_pseudobulks.csv.gz", row.names = 1)
common <- intersect(rownames(illumina_pseudobulks), rownames(ultima_pseudobulks))
union_id <- union(colnames(illumina_pseudobulks), colnames(ultima_pseudobulks))
combined_pseudobulks <- matrix(0,nrow = length(common), ncol = length(union_id), dimnames = list(common, union_id))
for (i in colnames(illumina_pseudobulks)) combined_pseudobulks[,i] <- illumina_pseudobulks[common, i]
for (i in colnames(ultima_pseudobulks)[!colnames(ultima_pseudobulks) %in% colnames(illumina_pseudobulks)]) combined_pseudobulks[,i] <- combined_pseudobulks[,i] + ultima_pseudobulks[common, i]

metadata <- read.csv("data_ageing/patient_level_metadata.csv")
rownames(metadata) <- paste0("WIS_blood_aging_", metadata$Individual.ID)
metadata_use<- metadata[colnames(combined_pseudobulks),]
S <- CreateSeuratObject(counts = combined_pseudobulks, names.field = 4, meta.data = metadata_use)
S <- NormalizeData(S, normalization.method = "LogNormalize", scale.factor = 10000)
S <- subset(S, cells = rownames(metadata))
S <- FindVariableFeatures(S, verbose=F)
S <- ScaleData(S, verbose=F)
S <- RunPCA(S, verbose=F)
metadata_final <- metadata[colnames(S),]

```

## Decompose data into GRPs

```{r decompose}
result <- decompose(s = S)
```

## Compute p-values: Association of GRPs with age (Figure S4A)

The `associate` function, in its basic use, computes p-value associations.

```{r pvalues, fig.width=5.5, fig.height=2.5}
pvalues <- associate_numeric_outcome(result, metadata_final, "age", run.ML = F)$p.values

pvalues$show_value <- with(pvalues, ifelse(p_outcome < 0.05, -log10(p_outcome), -log10(0.05)))
pvalues$factor_id <- factor(pvalues$factor_id, levels = sprintf("Factor_%d",USEGRPS))
ylabels <- gsub("Factor_", "", factor_annotation$factor_id); names(ylabels) <- factor_annotation$factor_id
any_sig <- sapply(names(ylabels), function(x) any(pvalues$p_outcome[pvalues$factor_id == x] < 0.06))
ylabels[!any_sig] <- ""
pvalues <- merge(pvalues, factor_annotation)

pvalues$xval <- with(pvalues, show_value * sign_outcome * multiply_with)
ylab <- sprintf("%s: %s", gsub("Factor_", "GRP", factor_annotation$factor_id), factor_annotation$factor_label); names(ylab) <- factor_annotation$factor_id

a <- qplot(x =  xval, y = reorder(factor_id,xval), color = as.factor(sign_outcome* multiply_with), data = subset(pvalues, p_outcome < 0.05)) +
  xlab("Association with age:\n-log10 p-value (signed)") + scale_y_discrete(labels = ylab, name = "GRP") + theme_bw(base_size = 14) + theme(panel.grid = element_blank()) + scale_color_manual(values = c("-1" = "darkblue", "1"= "darkred"), guide =F)
 print(a)
```

## Display correlation of GRP3 and Ageing (Figure 3B)

```{r simpleplot, fig.width=2.5, fig.height=2.5}
b <- qplot(result@result[ "Factor_3", ], metadata_final$age) +theme_bw(base_size = 14) + theme(panel.grid = element_blank()) + xlab("Factor 3") + ylab("Age")
print(b)
```

## Construct a GRP based ageing clock (Figure 3C)

First, run 10-fold CV with the factors to predict age

```{r grpclock}
factors <- rownames(result@result)[-1]
sets <- rep(1:10, length.out = ncol(S))
sh <- sample(1:ncol(S))
features <- t(result@result)[sh,-1]
response <- metadata_final[colnames(result), "age"][sh]
predicted.linear <- rep(NA, ncol(S)) 

for (i in 1:10) {
  train <- features[sets != i, ]
  test <- features[sets == i,]
  linear <- lm(response ~ . , data = data.frame(response = response[sets!=i], train))
  predicted.linear[sets == i] <- predict(linear, newdata = as.data.frame(test))
}
```

Then, run several machine learning strategies to predict age from all available transcriptomic variables

```{r associateML}
ml <- associate_numeric_outcome(result, metadata_final, "age", other.predictors = NULL, run.ML = T, replicates = 10, cores = 10, genelists = NULL)
```

Finally, plot

```{r plotClock, fig.width=3.5,fig.height=2.5}
r2s <- data.frame(
  what = factor(c("GRP 3", "All GRPs", "Full transcriptome"),levels = c("GRP 3", "All GRPs", "Full transcriptome")),
  R2 = c(cor(result@result[ "Factor_3", ], metadata_final$age)^2,cor(predicted.linear, response)^2, max(ml$r.squares, na.rm=T) )
)

d <- ggplot(aes(x = what, y = R2), data = r2s) + geom_col() + coord_flip()  +theme_bw(base_size = 14) + theme(panel.grid = element_blank()) + xlab("Predictor") + ylab("R2 in 10-fold CV")
print(d)

```

