---
title: "Decomposing LARRY data"
author: "Lars Velten"
date: "2025-08-15"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntryDeomposing LARRY data
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description of the data

Here, we decompose the transcriptomes of HSCs with known clonal behavior from [Rodriguez-Fraticelli et al., 2025](https://www.nature.com/articles/s41586-020-2503-6). Single-cell RNA-seq was performed on clonally barcoded cells post transplant. Clonal barcodes link HSCs to their output. Specifically, the paper defines the output of an HSC clone as the ratio between non-HSCs and HSCs per clone.

## Prepare data

We subset the data for HSCs, and compute clonal output as defined in the methods, section "Quantification and classification of HSC clonal behaviours" of [Rodriguez-Fraticelli et al., 2025](https://www.nature.com/articles/s41586-020-2503-6).

```{r load_packages, warning=F, message=F}
require(Seurat)
require(hsc18)
require(plyr)
require(ggplot2)
```

```{r prepate_data}
larry <- readRDS("/users/lvelten/lvelten/Data/Scherer2025/seurat_LARRY_RNAseq_LK_LSK.rds") #replace by figshare
larry$LARRY <- gsub("_","-", larry$LARRY)
#create HSC pseudobulks
larry_HSC <- subset(larry, CellType %in% c("HSC/MPP1")) 

larry_HSC_pseudobulk <- AggregateExpression(larry_HSC, group.by = "LARRY", return.seurat = T, verbose = F)
larry_HSC_pseudobulk <- FindVariableFeatures(larry_HSC_pseudobulk, verbose = F)
#compute clonal function
clone_info <- ddply(larry@meta.data, "LARRY",summarise,
                    n = length(LARRY),
                    n_hsc = sum(CellType == "HSC/MPP1"),
                    n_nonhsc = sum(CellType != "HSC/MPP1"))
clone_info <- within(clone_info,{
  f_clonesize <- n/sum(n)
  A_output <- (n_nonhsc / sum(n_nonhsc)) / (0.0001 + (n_hsc / sum(n_hsc)))
  use <- n_hsc > 0
})
```
## Decompose data

```{r run_decompose}
result <- decompose(s = larry_HSC_pseudobulk)
print(result)
```

## Associate with functional information

To link the processes with function, we need a data frame that contains for each RNA-seq sample, (in this case, clones) information on the function (Here: HSC clonal output). Row names need to match the colmn (sample) names of `result`.
To the metadata frame, we can also add any other gene signatures that we want to compare (in this case, gene lists from supplementary table 1 of )

```{r prepare_metadata, warning=F}
data("alejo_lists")
clone_info <- subset(clone_info, !is.na(LARRY))
rownames(clone_info) <- clone_info$LARRY
larry_HSC_pseudobulk <- AddModuleScore(larry_HSC_pseudobulk, features = alejo_lists, name = "alejo", verbose = F)
clone_info <- merge(clone_info, larry_HSC_pseudobulk@meta.data[, grepl("LARRY|alejo", colnames(larry_HSC_pseudobulk@meta.data))])
colnames(clone_info)[grepl("alejo", colnames(clone_info))] <- names(alejo_lists)

clone_info$output_log <- log10(clone_info$A_output+ 0.01)
rownames(clone_info) <- clone_info$LARRY
```

Equipped with this data frame, we can run `associate_numeric_outcome` (there is also `associate_binary_outcome` and `associate_survival_outcome`, but here we're looking at a numeric quantity - HSC output).

```{r associate}
grp2output <- associate_numeric_outcome(result, clone_info, "output_log", other.predictors = names(alejo_lists), run.pcr = T, replicates = 10, cores = 10, genelists = alejo_lists)
```
Let's plot the significances:

```{r pval.plot, fig.width=8,fig.height=4}
ggplot(aes(y = factor_id, x = -log10(p.adjust(p_outcome , method = "fdr"))),data = grp2output$p.values) +
         facet_wrap(~ifelse(grepl("Factor_", factor_id), "18 GRPs", "Other gene lists"), scales = "free_y") + 
         geom_point() +theme_bw() + xlab("-log10 FDR") + ylab("")
         
```


```{r rsqr.plot, fig.width=6,fig.height=6}
plf <- data.frame(what = names(grp2output$r.squares), value = unname(grp2output$r.squares), sd = grp2output$r.squares.sd)
plf$type <- with(plf, ifelse(what %in% c("R2.best.factor", "R2.best.other"  ), "0-shot",
                             ifelse(grepl("^R2", what), ifelse(grepl("factor", what), "ML+Factors","ML strategies"), "ML+Gene lists")))
ggplot(aes(y = reorder(what, value), x = value),data = subset(plf)) +
  geom_errorbarh(aes(xmin = value-sd, xmax = value+sd), height=0.1) +
         geom_point() +theme_bw() + xlab("R2 (10-fold CV)") +
  facet_grid(type~., space = "free_y", scales="free_y")
         
```


```{r rsqr.best, fig.width=4,fig.height=3}
best <- split(plf, plf$type)
best$`ML strategies` <- best$`ML strategies`[which.max(best$`ML strategies`$value),]
best$`ML+Gene lists` <- best$`ML+Gene lists`[which.max(best$`ML+Gene lists`$value),]
best$`ML+Factors` <- best$`ML+Factors`[which.max(best$`ML+Factors`$value),]

labels <- c("Best GRP", "Best signature", "Best ML recipe","lm of GRPs", "lm of signatures")
names(labels) <- c("R2.best.factor", "R2.best.other", best$`ML strategies`$what,best$`ML+Factors`$what, best$`ML+Gene lists`$what )

best <- do.call(rbind, best)


ggplot(aes(y = reorder(what, value), x = value, color = grepl("best", what)),data = subset(best)) +
  geom_errorbarh(aes(xmin = value-sd, xmax = value+sd), height=0.1) +
         geom_point() +theme_bw() + xlab("R2 (10-fold CV)") + scale_y_discrete(labels = labels) +
  scale_color_manual(name = "Class",labels = c("TRUE" = "0-shot", "FALSE" = "linear models"), values = c("TRUE" = "#CC0000", "FALSE" = "#333333"))
```
