---
title: "Make plots for SCEPTRE DE (Figure S2)"
author: "Joe Bowness"
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  rmd: "FigS2_CRISPRi500_sceptre"
abstract: To run the sceptre analyses see clean_analysis/sceptre. First I needed to make appropriate input files (from the CRISPRi500 seurat objects) with the script "make_sceptre_ondisc_objects_cluster_filterNT.R". Then actually run sceptre with "run_sceptre_cluster_ondisc_object.R". I later repeated the process to run differential expression on a per-guide basis (to approximately quantify consistency between guides). 
output_dir: "/users/lvelten/jbowness/CRISPRi500/HSC/clean_analysis/manuscript/"
output:
  html_document:
    number_sections: yes
    toc: yes
    code_folding: hide
    toc_depth: 4
    toc_float:
      collapsed: false
    highlight: pygments
editor_options:
  
  chunk_output_type: console
---
## Required packages

```{r, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(readr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(Seurat)
library(sceptre)
library(tidyr)
library(VennDiagram)
library(gridExtra)
library(ondisc)
library(plotly)
library(htmlwidgets)

```

# Load sceptre files

```{r}
## Load outputs from sceptre 
path_to_sceptre_objects <- "/users/lvelten/jbowness/CRISPRi500/HSC/clean_analysis/sceptre/sceptre_objects"
sceptre_files <- list.files(path_to_sceptre_objects, pattern = "\\.rds$", full.names = TRUE)
sceptre_files <- sceptre_files[
  sceptre_files != "/users/lvelten/jbowness/CRISPRi500/HSC/clean_analysis/sceptre/sceptre_objects/CRISPRi500_HSC_Rep1_Rem_rmNT12_byguide_default.rds"
]
sceptre_files <- sceptre_files[!grepl("Rep2", sceptre_files)]

for (f in sceptre_files) {
  nm <- tools::file_path_sans_ext(basename(f))
  e  <- new.env(parent = emptyenv())
  load(f, envir = e)
  message(nm)
  base::assign(x = nm, value = e[[ls(e)[1]]], envir = .GlobalEnv)
}
```

```{r}
## Load DE results (default settings)
path_to_sceptre_de_results <- "/users/lvelten/jbowness/CRISPRi500/HSC/clean_analysis/sceptre/de_tests"

Rep1_5X_sceptre_de <- read_csv(file.path(path_to_sceptre_de_results, "CRISPRi500_HSC_Rep1_5X_rmNT10_default.csv"))
Rep1_5X_sceptre_de <- Rep1_5X_sceptre_de %>% select(-"fold_change", -"se_fold_change")

Rep1_Rem_sceptre_de <- read_csv(file.path(path_to_sceptre_de_results, "CRISPRi500_HSC_Rep1_Rem_rmNT12_default.csv"))
Rep1_Rem_sceptre_de <- Rep1_Rem_sceptre_de %>% select(-"fold_change", -"se_fold_change")

# Rep 2 (only 5X genes)
Rep2_5X_sceptre_de <- read_csv(file.path(path_to_sceptre_de_results, "CRISPRi500_HSC_Rep2_5X_default.csv"))

```

# Sceptre plots

## Function to summarise num DE genes

```{r}
de_sum_cal_disc <- function(sceptre_obj, replicate, genes, covs, rmNT, cal_disc) {
  if(cal_disc == "cal"){
  #summarise number of significant DE genes per target
  df <- sceptre_obj@calibration_result %>%
    group_by(grna_target) %>%
    summarize(count = sum(significant == TRUE), 
              .groups = "keep")
  }
  else {
    df <- sceptre_obj@discovery_result %>%
      na.omit() %>%
      group_by(grna_target) %>%
      summarize(count = sum(significant == TRUE), 
                .groups = "keep")
    }
  #add additional description columns
  df$replicate <- replicate
  df$genes <- genes
  df$rmNT <- rmNT
  df$covs <- covs
  df$cal_disc <- cal_disc
  return(df)
}
```

## Add summaries of DE genes to larger output file

```{r}
de_sum_Rep1 <- data.frame(
  "grna_target" = character(), "count" = numeric(),
  "replicate" = character(), "genes" = character(), "rmNT" = character(),
  "covs" = character(), "cal_disc" = character(),
    stringsAsFactors = FALSE
)

for (f in sceptre_files) {
   # extract information from file name
   base <- tools::file_path_sans_ext(basename(f))
   message(base)
   m <- regexec("^CRISPRi500_HSC_(Rep\\d+)_([A-Za-z0-9]+)(?:_(rmNT\\d+))?_(default|\\dPC)$", base)
   g <- regmatches(base, m)[[1]]
   rep   <- g[2]                                    # e.g., "Rep1"
   rem5X <- g[3]                                  # "5X" or "Rem"
   rmNT  <- if (length(g) >= 4 && !is.na(g[4]) && nzchar(g[4])) g[4] else NA_character_  # e.g., "rmNT10"
   covs <- g[5]                                # "default" | "2PC" | "3PC"
   
   # get X5 instead of 5X
   remX5 <- if (rem5X == "5X") "X5" else rem5X
   # get default, PC2 or PC3
   if (covs == "default") {
      covs <- "def"
    } else {
      covs <- sub("^(\\d)PC$", "PC\\1", covs)
    }
   
   obj <- get(base, envir = .GlobalEnv)
   
   # collect calibration results
   out <- de_sum_cal_disc(sceptre_obj = obj, 
                   replicate = rep, 
                   genes = rem5X, 
                   covs = covs, 
                   rmNT = rmNT, 
                   cal_disc = "cal")
   
   # (OPTIONAL) save to separate output file
    #var_nm <- paste0(
    #  "de_sum_", rem5X, "_", rep, "_", covs,
    #  if (!is.na(rmNT)) paste0("_", rmNT),
    #  "_cal")
   #assign(var_nm, out, envir = .GlobalEnv)
   
   # add to overall results table
   de_sum_Rep1 <- rbind(de_sum_Rep1, out)
   
   # collect discovery results and add to results
   out <- de_sum_cal_disc(sceptre_obj = obj, 
                   replicate = rep, 
                   genes = rem5X, 
                   covs = covs, 
                   rmNT = rmNT, 
                   cal_disc = "disc")
   de_sum_Rep1 <- rbind(de_sum_Rep1, out)
   }

## add hasNT column
de_sum_Rep1 <- de_sum_Rep1 %>%
  mutate(has_NT = grepl("NO-SITE-sgRNA-12", grna_target, fixed = TRUE) | 
           grepl("NO-SITE-sgRNA-10", grna_target, fixed = TRUE))

```

## Plot violin plots of false discoveries 

```{r}
false_pos_violins <- subset(de_sum_Rep1, cal_disc == "cal" & replicate == "Rep1") %>%
  ggplot(aes(x = covs, y = count, fill = covs)) +
  geom_violin(trim = TRUE, alpha = 0.6) +
  geom_boxplot(width = 0.1, position = position_dodge(width = 0.8),  alpha = 0.8) +  
  geom_jitter(aes(color = has_NT), width = 0, shape = 16, size = 2, alpha = 0.6) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "darkgrey")) +
  labs(  x = "covariate model",
    y = "number of false positives per NT group",
    color = "Group contains NT10 or NT12") +
   guides(fill = "none") +
  scale_y_continuous(trans = 'log1p', breaks = c(0, 1, 5, 10, 50, 100, 150))+
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  facet_wrap(~ genes + rmNT, nrow = 2)
false_pos_violins
```

### only default settings (no covariates)

```{r}
de_sum_Rep1_filtered <- subset(de_sum_Rep1, cal_disc == "cal" & covs == "def" & replicate == "Rep1")
de_sum_Rep1_filtered <- de_sum_Rep1_filtered %>%
  mutate(
    rmNT = case_when(
      rmNT %in% c("rmNT10", "rmNT12") ~ "rmNT",
      is.na(rmNT) ~ "incNT",
      TRUE ~ rmNT
    )
  )

false_pos_violin_def <- ggplot(de_sum_Rep1_filtered, aes(x = genes, y = count, fill = rmNT)) +
  geom_violin(trim = TRUE, alpha = 0.6) +
  geom_boxplot(width = 0.1, position = position_dodge(width = 0.8),  alpha = 0.8) +  
  geom_jitter(aes(color = has_NT), width = 0, shape = 16, size = 2, alpha = 0.6) +
  scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "darkgrey")) +
  labs(  x = "include or remove NT10/N12",
    y = "number of false positives per NT group",
    color = "Group contains NT10 or NT12") +
   guides(fill = "none") +
  scale_y_continuous(trans = 'log1p', breaks = c(0, 1, 5, 10, 50, 100, 150))+
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  facet_wrap(~ rmNT, nrow = 1, drop=TRUE)
false_pos_violin_def

```

```{r, eval=FALSE}
# save plot
ggsave( filename = "manuscript/figs/sceptre_calibration_violin_false_pos.pdf",
        plot     = false_pos_violin_def,
        width    = 8,
        height   = 5,
        units    = "in"
        )
```


## Plot of total DE genes

```{r}
# summarise total DE genes
de_sum_Rep1_total <- de_sum_Rep1 %>%
  group_by(genes, rmNT, covs, cal_disc) %>%
  summarize(total_DE = sum(count), .groups = "drop")

# Set factors so NAs on the left
de_sum_Rep1_total$rmNT <- as.character(de_sum_Rep1_total$rmNT)   
de_sum_Rep1_total$rmNT[is.na(de_sum_Rep1_total$rmNT)] <- "NA" 
de_sum_Rep1_total$rmNT <- factor(de_sum_Rep1_total$rmNT,
                                 levels = c("NA", "rmNT10", "rmNT12"))

## split X5 and Rem
df_X5  <- filter(de_sum_Rep1_total, genes == "5X")
df_Rem <- filter(de_sum_Rep1_total, genes == "Rem")

## plotting function
plot_by_genes <- function(df, title = NULL) {
  ggplot(df, aes(x = covs, y = total_DE, fill = cal_disc)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ rmNT, scales = "fixed", drop = TRUE) +
    labs(title = title, x = "Covariates", y = "Total DE genes") +
    theme_minimal()
}

## make plots
p_X5  <- plot_by_genes(df_X5,  "Total DE for targets at 5X ")
p_Rem <- plot_by_genes(df_Rem, "Total DE for remaining targets")

p_X5
p_Rem
```

### total DE genes for only default settings 

```{r}
## plotting function
plot_by_genes_def <- function(df, title = NULL) {
  df %>% subset(covs == "def") %>%
  ggplot(aes(x = covs, y = total_DE, fill = cal_disc)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(
      aes(label = total_DE),
      position = position_dodge(width = 0.9),
      vjust = -0.3,
      size = 3) +
    facet_wrap(~ rmNT, scales = "fixed", drop = TRUE) +
    labs(title = title, x = "Covariates", y = "Total DE genes") +
    theme_minimal()
}

## make plots
p_X5_def  <- plot_by_genes_def(df_X5,  "Total DE for targets at 5X ")
p_Rem_def <- plot_by_genes_def(df_Rem, "Total DE for remaining targets")

p_X5_def + p_Rem_def
```

```{r, eval=FALSE}
# save plot
ggsave( filename = "manuscript/figs/sceptre_calibration_bar_totalDE_5X.pdf",
        plot     = p_X5_def,
        width    = 4,
        height   = 5,
        units    = "in"
        )
ggsave( filename = "manuscript/figs/sceptre_calibration_bar_totalDE_Rem.pdf",
        plot     = p_Rem_def,
        width    = 4,
        height   = 5,
        units    = "in"
        )
```

## How does cell number affect num of DE genes?

```{r}
## get cells per target gene for 5X genes ####
obj <- CRISPRi500_HSC_Rep1_5X_rmNT10_default
idxs <- obj@grna_assignments[["grna_group_idxs"]]
num_cells <- vapply(idxs, length, integer(1))

cells_per_target_5X <- data.frame(
  grna_target = names(num_cells),
  num_cells = unname(num_cells),
  row.names = NULL,
  stringsAsFactors = FALSE
)

# add this to the DE genes summary
DE_per_target_5X <- subset(de_sum_Rep1,
                        replicate == "Rep1" &
                          genes == "5X" &
                          !(is.na(rmNT)) &
                          covs == "def" &
                          cal_disc == "disc")
DE_per_target_5X <- merge(DE_per_target_5X, cells_per_target_5X, by="grna_target")

## get cells per target gene for Remaining genes ####
obj <- CRISPRi500_HSC_Rep1_Rem_rmNT12_default
idxs <- obj@grna_assignments[["grna_group_idxs"]]
num_cells <- vapply(idxs, length, integer(1))

cells_per_target_Rem <- data.frame(
  grna_target = names(num_cells),
  num_cells = unname(num_cells),
  row.names = NULL,
  stringsAsFactors = FALSE
)

# add this to the DE genes summary
DE_per_target_Rem <- subset(de_sum_Rep1,
                        replicate == "Rep1" &
                          genes == "Rem" &
                          !(is.na(rmNT)) &
                          covs == "def" &
                          cal_disc == "disc")
DE_per_target_Rem <- merge(DE_per_target_Rem, cells_per_target_Rem, by="grna_target")

## merge together and plot ####
DE_per_target <- rbind(DE_per_target_5X, DE_per_target_Rem)

# Compute Pearson correlation
cor_val <- cor(DE_per_target$num_cells,
               DE_per_target$count,
               method = "pearson",
               use = "complete.obs")

scatter_ncells_DE <- ggplot(DE_per_target, 
                            aes(x = num_cells, y = count)) +
  geom_point(size = 2, alpha = 0.7) +
  #geom_smooth(method = "lm", linewidth = 0.5, se = FALSE) +  
  scale_x_continuous(trans = "log1p",
                     breaks = c(0, 1, 5, 10, 20, 50, 100, 500, 1000),
                     limits = c(1, NA),
                     expand = c(0.1, 0)) + 
  scale_y_continuous(trans = "log1p",
                     breaks = c(0, 1, 5, 10, 20, 50, 100, 500, 1000),
                     limits = c(1, NA),
                     expand = c(0.1, 0)) +
  labs(x = "Number of cells", y = "Number of DE genes discovered") +
  theme_minimal() +
  annotate("text",
           x = Inf, y = Inf,
           hjust = 1.1, vjust = 1.5,
           label = paste0("R = ", round(cor_val, 2)),
           size = 5) +
  theme_minimal()

scatter_ncells_DE

```

```{r, eval=FALSE}
# save plot
ggsave( filename = "manuscript/figs/scatter_ncells_DE.pdf",
        plot     = scatter_ncells_DE,
        width    = 6,
        height   = 5,
        units    = "in"
        )
```

## Add consistency score to DE table

```{r}
## load 5X genes sceptre results (testing each guide RNA separately)
CRISPRi500_HSC_Rep1_5X_rmNT10_byguide_default <- read_csv("/users/lvelten/jbowness/CRISPRi500/HSC/clean_analysis/sceptre/de_tests/CRISPRi500_HSC_Rep1_5X_rmNT10_byguide_default.csv")
CRISPRi500_HSC_Rep1_5X_rmNT10_byguide_default <- CRISPRi500_HSC_Rep1_5X_rmNT10_byguide_default %>%
  mutate(
    grna_target = as.character(grna_target),
    target_gene = sub("-CRISPRi-sgRNA-[0-9]+$", "", grna_target)
  )

# function to summarise guide consistency
summarise_crispri_consistency <- function(df) {
  eps_low  <- 1e-300
  eps_high <- 1e-16
  
  # ---- base summary per (response_id, grna_target) ----
  base_summary <- na.omit(df) %>%
    dplyr::group_by(response_id, target_gene) %>%
    dplyr::summarise(
      n_guides   = sum(!is.na(log_2_fold_change)),
      mean_log2FC = mean(log_2_fold_change, na.rm = TRUE),
      sd_log2FC   = sd(log_2_fold_change,   na.rm = TRUE),
      sign_agreement = {
        m <- mean(log_2_fold_change, na.rm = TRUE)
        mean(sign(log_2_fold_change) == sign(m), na.rm = TRUE)
      },
      
      mean_neglog10p = mean(-log10(pmax(p_value, .Machine$double.xmin)), na.rm = TRUE),
      
      # ---- Stouffer's Z meta-analysis ----
      z = {
        p_clamped <- pmin(pmax(p_value, eps_low), 1 - eps_high)
        z_i <- sign(log_2_fold_change) * qnorm(1 - p_clamped / 2)
        z_i <- pmax(pmin(z_i, 8), -8)  # optional winsorization
        sum(z_i, na.rm = TRUE) / sqrt(sum(!is.na(z_i)))
      },
      p_meta = 2 * (1 - stats::pnorm(abs(z))),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      # binomial weighting of sign agreement sqrt(n_guides)
      sd_log2FC = ifelse(is.na(sd_log2FC), 0, sd_log2FC),
      consistency_score = (sign_agreement * sqrt(n_guides)) * mean_neglog10p / (1 + sd_log2FC) 
    ) %>%
    dplyr::mutate(
      p_meta_adj = p.adjust(p_meta, method = "BH")
    )

  
  # ---- extract guide index from grna_target and pivot to wide ----
  wide_guides <- df %>%
    dplyr::mutate(
      grna_target = as.character(grna_target),
      guide_num = suppressWarnings(as.integer(sub(".*-CRISPRi-sgRNA-","", grna_target)))
    ) %>%
    dplyr::filter(!is.na(guide_num), guide_num %in% 1:3) %>%
    dplyr::group_by(response_id, target_gene, guide_num) %>%
    dplyr::summarise(
      log2FC = dplyr::first(log_2_fold_change),
      p     = dplyr::first(p_value),
      n_nonzero_trt = dplyr::first(n_nonzero_trt),
      .groups = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from  = guide_num,
      values_from = c(log2FC, p, n_nonzero_trt),
      names_glue  = "{.value}_g{guide_num}"
    )

  # ---- final join ----
  result <- base_summary %>%
    dplyr::left_join(wide_guides, by = c("response_id","target_gene"))
  
  return(result)
}


consistency_summary <- summarise_crispri_consistency(
  CRISPRi500_HSC_Rep1_5X_rmNT10_byguide_default
)

Rep1_5X_sceptre_de_ext <- Rep1_5X_sceptre_de %>%
  rename("target_gene" = "grna_target") %>%
    dplyr::left_join(consistency_summary, by = c("response_id","target_gene"))

write_csv(Rep1_5X_sceptre_de_ext, 
          "/users/lvelten/jbowness/CRISPRi500/HSC/clean_analysis/manuscript/int_files/Rep1_5X_sceptre_de_ext.csv")

```

## Repeat for remaining genes

```{r, eval=FALSE}
# Load DE from remaining genes
CRISPRi500_HSC_Rep1_Rem_rmNT12_byguide_default <- read_csv("/users/lvelten/jbowness/CRISPRi500/HSC/clean_analysis/sceptre/de_tests/CRISPRi500_HSC_Rep1_Rem_rmNT12_byguide_default.csv")
CRISPRi500_HSC_Rep1_Rem_rmNT12_byguide_default <- CRISPRi500_HSC_Rep1_Rem_rmNT12_byguide_default %>%
  mutate(
    grna_target = as.character(grna_target),
    target_gene = sub("-CRISPRi-sgRNA-[0-9]+$", "", grna_target)
  )

consistency_summary_rem <- summarise_crispri_consistency(
  CRISPRi500_HSC_Rep1_Rem_rmNT12_byguide_default
)

Rep1_Rem_sceptre_de_ext <- Rep1_Rem_sceptre_de %>%
  rename("target_gene" = "grna_target") %>%
    dplyr::left_join(consistency_summary_rem, by = c("response_id","target_gene"))

write_csv(Rep1_Rem_sceptre_de_ext, 
          "/users/lvelten/jbowness/CRISPRi500/HSC/clean_analysis/manuscript/int_files/Rep1_Rem_sceptre_de_ext.csv")

```

# Volcano plots for some example genes

```{r}
# function to plot volcano plot
plot.volcano <- function(de_df, gene, label_pval, label_log2,
                         alpha_min = 0.1, alpha_max = 1.0,
                         meta_thresh = 0.05) {

  # safe -log10(p)
  de_df$neg_log10_pval <- pmin(-log10(pmax(de_df$p_value, .Machine$double.xmin)), 10)

  # subset
  de_df_gene <- subset(de_df, target_gene == gene & pass_qc == "TRUE")

  # labels
  de_df_gene$label <- ifelse(
    de_df_gene$neg_log10_pval > label_pval |
      abs(de_df_gene$log_2_fold_change) > label_log2,
    de_df_gene$response_id, ""
  )
  ## --- scaling helper ---
  rescale_safe <- function(x, out_min, out_max) {
    if (all(is.na(x))) return(rep(out_min, length(x)))
    x[!is.finite(x)] <- NA
    rng <- range(x, na.rm = TRUE)
    if (diff(rng) == 0) return(rep(out_max, length(x)))
    x_scaled <- (x - rng[1]) / diff(rng)
    x_scaled[is.na(x_scaled)] <- 0
    out_min + x_scaled * (out_max - out_min)
  }
  
  # scale alpha by |z|
  de_df_gene$alpha_pt <- rescale_safe(abs(de_df_gene$z), alpha_min, alpha_max)

  # robust logical for 'significant' (handles TRUE/FALSE or "TRUE"/"FALSE")
  sig <- if (is.logical(de_df_gene$significant)) {
    de_df_gene$significant
  } else {
    toupper(as.character(de_df_gene$significant)) == "TRUE"
  }

  # color grouping
  de_df_gene$color_grp <- dplyr::case_when(
    !sig ~ "ns",  # not significant
    sig & de_df_gene$log_2_fold_change > 0 & de_df_gene$p_meta_adj <= meta_thresh ~ "up_strong",
    sig & de_df_gene$log_2_fold_change > 0 & de_df_gene$p_meta_adj > meta_thresh ~ "up_weak",
    sig & de_df_gene$log_2_fold_change < 0 & de_df_gene$p_meta_adj <= meta_thresh ~ "down_strong",
    sig & de_df_gene$log_2_fold_change < 0 & de_df_gene$p_meta_adj > meta_thresh ~ "down_weak",
    TRUE ~ "ns"
  )
  
  # palette
  cols <- c(
    ns          = "grey70",
    up_weak     = "#e88484",  # light red
    up_strong   = "#b80707",  # dark red
    down_weak   = "#42a1f5",  # light blue
    down_strong = "#0D47A1"   # dark blue
  )

  plot_title <- paste("Volcano plot -", gene)

  p <- ggplot(de_df_gene, aes(x = log_2_fold_change, y = neg_log10_pval)) +
    geom_point(
      aes(text = response_id, color = color_grp, alpha = alpha_pt)
    ) +
    scale_color_manual(values = cols, name = "Significance") +
    scale_alpha_identity() +
    geom_text(aes(label = label), vjust = -0.5, hjust = 0.5, size = 3, check_overlap = FALSE) +
    labs(x = "Log2 Fold Change", y = "-log10(P-value)", title = plot_title) +
    theme_minimal()

  return(p)
}

# plot volcanos
#Msi2_volcano <- plot.volcano(Rep1_Rem_sceptre_de_ext, "Msi2", 4, 1.4)
#Msi2_volcano

Runx1_volcano <- plot.volcano(Rep1_5X_sceptre_de_ext, "Runx1", label_pval = 6, label_log2 = 1.2)
Runx1_volcano

# check overlap between significance and p_meta_adj
Rep1_5X_sceptre_de_ext %>%
  subset(target_gene == "Runx1") %>%
  summarise(
    n_total      = n(),
    n_signif     = sum(significant == TRUE, na.rm = TRUE),
    n_meta       = sum(p_meta_adj < 0.05, na.rm = TRUE),
    n_overlap    = sum(significant == TRUE & p_meta_adj < 0.05, na.rm = TRUE),
    prop_overlap = n_overlap / n_signif
    )
```

## plotly volcano

```{r}
plot.volcano.plotly <- function(de_df, gene,
                                meta_thresh = 0.05) {
  stopifnot(all(c("target_gene","pass_qc","p_value","response_id",
                  "log_2_fold_change","significant","p_meta_adj",
                  "consistency_score","z") %in% names(de_df)))

  # safe -log10(p)
  de_df <- de_df %>%
    mutate(neg_log10_pval = pmin(-log10(pmax(p_value, .Machine$double.xmin)), 10))

  # subset to this gene and QC-passing rows
  de_df_gene <- subset(de_df, target_gene == gene & pass_qc == "TRUE")

  # scale alpha by |z|
  alpha_from_z <- function(z, alpha_min = 0.05, alpha_max = 1) {
  z_abs <- abs(z)
  z_abs[!is.finite(z_abs)] <- NA

  # logistic parameters: shape tuned to hit ~|z|=1→0.2, |z|=2→0.6, |z|=10→1
  k <- 0.9   # shallower slope (gentler rise)
  c <- 3.5   # midpoint around z = 2
  scaled <- 1 / (1 + exp(-k * (z_abs - c)))

  # normalize so that |z| = 10 saturates at 1
  scaled <- pmin(scaled / (1 / (1 + exp(-k * (10 - c)))), 1)

  scaled[is.na(scaled)] <- 0
  alpha_min + scaled * (alpha_max - alpha_min)
  }
  de_df_gene$alpha_pt <- alpha_from_z(de_df_gene$z)


  # robust logical for 'significant'
  sig <- if (is.logical(de_df_gene$significant)) de_df_gene$significant
         else toupper(as.character(de_df_gene$significant)) == "TRUE"

  # color groups (keep ns grey)
  de_df_gene$color_grp <- dplyr::case_when(
    !sig ~ "ns",
    sig & de_df_gene$log_2_fold_change > 0 & de_df_gene$p_meta_adj <= meta_thresh ~ "up_strong",
    sig & de_df_gene$log_2_fold_change > 0 & de_df_gene$p_meta_adj >  meta_thresh ~ "up_weak",
    sig & de_df_gene$log_2_fold_change < 0 & de_df_gene$p_meta_adj <= meta_thresh ~ "down_strong",
    sig & de_df_gene$log_2_fold_change < 0 & de_df_gene$p_meta_adj >  meta_thresh ~ "down_weak",
    TRUE ~ "ns"
  )

  # base hex colors per group
  base_cols <- c(
    ns          = "#B3B3B3",  # grey70
    up_weak     = "#e88484",  # light red
    up_strong   = "#b80707",  # dark red
    down_weak   = "#42a1f5",  # light blue
    down_strong = "#0D47A1"   # dark blue
  )

  # helper: hex -> rgba with per-point alpha
  hex_to_rgba <- function(hex, alpha_vec) {
    # strip '#', parse
    hex <- gsub("#","", hex)
    r <- strtoi(substr(hex,1,2),16L)
    g <- strtoi(substr(hex,3,4),16L)
    b <- strtoi(substr(hex,5,6),16L)
    paste0("rgba(", r, ",", g, ",", b, ",", pmin(pmax(alpha_vec, 0), 1), ")")
  }

  # ensure ordered legend
  grp_levels <- c("ns","up_weak","up_strong","down_weak","down_strong")
  de_df_gene$color_grp <- factor(de_df_gene$color_grp, levels = grp_levels)

  plt <- plot_ly()
  for (grp in grp_levels) {
    d <- de_df_gene %>% filter(color_grp == grp)
    if (nrow(d) == 0) next

    # rgba colors with individual alpha
    col_rgba <- hex_to_rgba(base_cols[[grp]], d$alpha_pt)
    
    # collapse per-guide logFCs, comma-separated
    fc_str <- apply(d[, c("log2FC_g1","log2FC_g2","log2FC_g3")], 1, function(x) {
      paste(na.omit(signif(x, 3)), collapse = ", ")
    })

    # tooltip:
    hover_txt <- paste0(
     "<b>", d$response_id, "</b> ",
      "<br>log2FC: ", signif(d$log_2_fold_change, 3),
      "<br>n_guides_pass_qc: ", d$n_guides,
      "<br>log2FC_guides: ", fc_str,
      "<br>consistency_score: ", signif(d$consistency_score, 3)   )

    plt <- add_trace(
      plt,
      type = "scattergl",
      mode = "markers",
      data = d,
      x = ~log_2_fold_change,
      y = ~neg_log10_pval,
      name = grp,
      text = hover_txt,
      hoverinfo = "text",
      marker = list(
        color = col_rgba,
        size  = 8,          # tweak if you want larger/smaller points
        line  = list(width = 0)
      ),
      showlegend = TRUE
    )
  }

  plot_title <- paste("Volcano plot -", gene)

  plt %>%
    layout(
      title = list(text = plot_title),
      xaxis = list(title = "Log2 Fold Change", zeroline = FALSE),
      yaxis = list(title = "-log10(P-value)", zeroline = FALSE),
      legend = list(title = list(text = NULL))
    )
}
plot.volcano.plotly(Rep1_5X_sceptre_de_ext, "Runx1")
```

## loop and save volcano plots

```{r, eval=FALSE}
# create folder if not already present
out_dir <- "manuscript/de_volcanos_5X"
if (!dir.exists(out_dir)) dir.create(out_dir)

# get all unique target genes
genes <- unique(Rep1_5X_sceptre_de_ext$target_gene)

# loop through and save one volcano plot per gene
for (g in genes) {
  cat("Saving volcano for:", g, "\n")
  p <- plot.volcano.plotly(de_df = Rep1_5X_sceptre_de_ext, gene = g)
  
  # shrink dependencies to only what's needed (e.g., scattergl)
  p <- plotly::partial_bundle(p)
  
  safe_name <- gsub("[^A-Za-z0-9_\\-]", "_", g)
  file_path <- file.path(out_dir, paste0("de_volcano_", safe_name, ".html"))

  #  build .html filename
  tmpfile <- tempfile(fileext = ".html")
  htmlwidgets::saveWidget(p, file = tmpfile, selfcontained = TRUE, libdir = NULL)

  # copy the fully inlined html into your final path
  txt <- readChar(tmpfile, file.info(tmpfile)$size, useBytes = TRUE)
  writeLines(txt, con = file_path, useBytes = TRUE)
}

```

## loop and save volcanos for remaining genes

```{r, eval=FALSE}
# create folder if not already present
out_dir <- "manuscript/de_volcanos_remaining"
if (!dir.exists(out_dir)) dir.create(out_dir)

# get all unique target genes
genes <- unique(Rep1_Rem_sceptre_de_ext$target_gene)

# loop through and save one volcano plot per gene
for (g in genes) {
  cat("Saving volcano for:", g, "\n")
  p <- plot.volcano.plotly(de_df = Rep1_Rem_sceptre_de_ext, gene = g)
  
  # shrink dependencies to only what's needed (e.g., scattergl)
  p <- plotly::partial_bundle(p)
  
  safe_name <- gsub("[^A-Za-z0-9_\\-]", "_", g)
  file_path <- file.path(out_dir, paste0("de_volcano_", safe_name, ".html"))

  #  build .html filename
  tmpfile <- tempfile(fileext = ".html")
  htmlwidgets::saveWidget(p, file = tmpfile, selfcontained = TRUE, libdir = NULL)

  # copy the fully inlined html into your final path
  txt <- readChar(tmpfile, file.info(tmpfile)$size, useBytes = TRUE)
  writeLines(txt, con = file_path, useBytes = TRUE)
}
```

# Concordance with replicate 2

```{r}
# test what happens to overlap coefficient with 2 or 3 PCs in sceptre model
## 2PCs
#Rep1_5X_sceptre_de <- read_csv(file.path(path_to_sceptre_de_results, "CRISPRi500_HSC_Rep1_5X_rmNT10_2PC.csv"))
#Rep2_5X_sceptre_de <- read_csv(file.path(path_to_sceptre_de_results, "CRISPRi500_HSC_Rep2_5X_2PC.csv"))
## 3 PCS
#Rep1_5X_sceptre_de <- read_csv(file.path(path_to_sceptre_de_results, "CRISPRi500_HSC_Rep1_5X_rmNT10_3PC.csv"))
#Rep2_5X_sceptre_de <- read_csv(file.path(path_to_sceptre_de_results, "CRISPRi500_HSC_Rep2_5X_3PC.csv"))

```

```{r}
#subset DE results dfs to the same grna_target and reponse_id
Rep1_5X_sceptre_de <- subset(Rep1_5X_sceptre_de, 
                          Rep1_5X_sceptre_de$response_id %in% Rep2_5X_sceptre_de$response_id &
                            Rep1_5X_sceptre_de$grna_target %in% Rep2_5X_sceptre_de$grna_target) 
Rep2_5X_sceptre_de <- subset(Rep2_5X_sceptre_de, 
                          Rep2_5X_sceptre_de$response_id %in% Rep1_5X_sceptre_de$response_id &
                            Rep2_5X_sceptre_de$grna_target %in% Rep1_5X_sceptre_de$grna_target) 

Rep1_5X_sceptre_de$replicate <- "Rep1"
Rep2_5X_sceptre_de$replicate <- "Rep2"
sceptre_5X_combined <- rbind(Rep1_5X_sceptre_de, Rep2_5X_sceptre_de)


#plots for individual perturbations
plot_replicate_agreement <- function(sceptre_obj, gene){
  gene_sceptre <- subset(sceptre_obj, grna_target == gene)
  # 1) Prepare data
  plot_df <- gene_sceptre %>%
  select(response_id, replicate, log2 = log_2_fold_change, significant) %>%
  filter(!is.na(log2)) %>%
  pivot_wider(
    names_from  = replicate,
    values_from = c(log2, significant),
    names_sep   = "_"
  ) %>%
  filter(!is.na(log2_Rep1) & !is.na(log2_Rep2)) %>%
  mutate(
    category = case_when(
      significant_Rep1 & significant_Rep2        ~ "both",
      xor(significant_Rep1, significant_Rep2)    ~ "one",
      TRUE                                        ~ "none"
    ),
    color_group = case_when(
      # both sig & both positive
      category=="both" & log2_Rep1>0  & log2_Rep2>0  ~ "both_pos",
      # one sig & that sig rep is positive
      category=="one"  & ((significant_Rep1 & !significant_Rep2 & log2_Rep1>0) |
                          (significant_Rep2 & !significant_Rep1 & log2_Rep2>0)) ~ "one_pos",
      # both sig & both negative
      category=="both" & log2_Rep1<0  & log2_Rep2<0  ~ "both_neg",
      # one sig & that sig rep is negative
      category=="one"  & ((significant_Rep1 & !significant_Rep2 & log2_Rep1<0) |
                          (significant_Rep2 & !significant_Rep1 & log2_Rep2<0)) ~ "one_neg",
      # everything else
      TRUE                                        ~ "other"
    )
  )
  # 2) Pearson correlation for annotation
  ct         <- cor.test(plot_df$log2_Rep1, plot_df$log2_Rep2, method="pearson", exact=FALSE)
  r_val      <- signif(ct$estimate, 3)
  corr_label <- paste0("Pearson R = ", r_val)
  # 3) Scatter plot with layered coloring
  scatter_plot <- ggplot() +
  # "other" (not sig in either OR mixed-sign)
  geom_point(data = filter(plot_df, color_group=="other"),
             aes(x = log2_Rep1, y = log2_Rep2),
             color = "grey70", alpha = 0.6) +
  # one sig, negative
  geom_point(data = filter(plot_df, color_group=="one_neg"),
             aes(x = log2_Rep1, y = log2_Rep2),
             color = "skyblue", alpha = 0.8) +
  # both sig, negative
  geom_point(data = filter(plot_df, color_group=="both_neg"),
             aes(x = log2_Rep1, y = log2_Rep2),
             color = "blue2", alpha = 0.9) +
  # one sig, positive
  geom_point(data = filter(plot_df, color_group=="one_pos"),
             aes(x = log2_Rep1, y = log2_Rep2),
             color = "firebrick2", alpha = 0.8) +
  # both sig, positive
  geom_point(data = filter(plot_df, color_group=="both_pos"),
             aes(x = log2_Rep1, y = log2_Rep2),
             color = "firebrick4", alpha = 0.9) +
  # identity line
  geom_abline(slope = 1, intercept = 0, color = "red") +
  coord_equal() +
  labs(x     = "Rep1 log2FC",
       y     = "Rep2 log2FC"  ) +
    ggtitle(paste(gene,": replicate comparison")) +
  theme_minimal()
  
  ##### PLOT VENN DIAGRAMS
  # 1) Compute sets of DE genes for Rep1 and Rep2, separately for up/down
  sig1_up   <- gene_sceptre %>%
    filter(replicate=="Rep1", significant, log_2_fold_change > 0) %>%
    pull(response_id) %>% unique()
  sig2_up   <- gene_sceptre %>%
    filter(replicate=="Rep2", significant, log_2_fold_change > 0) %>%
    pull(response_id) %>% unique()
  sig1_down <- gene_sceptre %>%
    filter(replicate=="Rep1", significant, log_2_fold_change < 0) %>%
    pull(response_id) %>% unique()
  sig2_down <- gene_sceptre %>%
    filter(replicate=="Rep2", significant, log_2_fold_change < 0) %>%
    pull(response_id) %>% unique()
  
  # make venn diagrams
  venn_up <- venn.diagram(
    x = list(Rep1 = sig1_up, Rep2 = sig2_up),
    filename = NULL,
    main     = "DE up",
    fill     = c("firebrick","firebrick1"),
    alpha    = 0.5,
    scaled   = TRUE,
    margin   = 0.05,
    cat.pos  = c(-20, 20),
    cat.dist = c(0.03, 0.03)
  )
  
  venn_down <- venn.diagram(
    x = list(Rep1 = sig1_down, Rep2 = sig2_down),
    filename = NULL,
    main     = "DE down",
    fill     = c("steelblue","skyblue"),
    alpha    = 0.5,
    scaled   = TRUE,
    margin   = 0.05,
    cat.pos  = c(-20, 20),
    cat.dist = c(0.03, 0.03)
  )
  
  # plot them as a grid
  lay <- rbind(c(1,2),
               c(1,3))
  grid.newpage()
  up_grob   <- grobTree(venn_up)
  down_grob <- grobTree(venn_down)
  plot_grid <- grid.arrange(
    scatter_plot,
    up_grob,
    down_grob,
    layout_matrix = lay,
    widths = c(2,1),
    heights = c(1,1)
  )
  return(plot_grid)
}
Spi1_rep_agreement <- plot_replicate_agreement(sceptre_5X_combined, "Spi1")
Spi1_rep_agreement
```

```{r, eval=FALSE}
# save plot
ggsave( filename = "manuscript/figs/Spi1_rep_agreement.pdf",
        plot     = Spi1_rep_agreement,
        width    = 7.5,
        height   = 5,
        units    = "in"
        )
```


## Overall overlap with Rep2

The **Overlap Coefficient**, also known as the Szymkiewicz–Simpson coefficient, is defined as the size of the union of set A and set B over the size of the smaller set between A and B.

In the case of default sceptre settings it is (368 + 445) / (368 + 445 + 216 + 106) = 0.716. It is actually less with 2PCs (0.68) or 3PCs (0.69) as covariates in sceptre model. 

```{r}
# How many significant genes from Rep2 are found in Rep1?
df <- sceptre_5X_combined

rep1_sig <- df %>%
  filter(replicate == "Rep1", significant) %>%
  select(grna_target, response_id, log_2_fold_change) %>%
  rename(log2FC_Rep1 = log_2_fold_change)

rep2_sig <- df %>%
  filter(replicate == "Rep2", significant) %>%
  select(grna_target, response_id, log_2_fold_change) %>%
  rename(log2FC_Rep2 = log_2_fold_change)

merged <- merge(rep2_sig, rep1_sig, by = c("grna_target","response_id"), all=TRUE) 
#build unqiue keys for up and down
df <- merged
df_ids <- df %>%
  filter(!is.na(grna_target), !is.na(response_id)) %>%
  mutate(pair_id = as.integer(interaction(grna_target, response_id, drop = TRUE)))

# --- build sets ---
rep1_up   <- df_ids %>% filter(!is.na(log2FC_Rep1), log2FC_Rep1 > 0) %>% pull(pair_id) %>% unique()
rep1_down <- df_ids %>% filter(!is.na(log2FC_Rep1), log2FC_Rep1 < 0) %>% pull(pair_id) %>% unique()
rep2_up   <- df_ids %>% filter(!is.na(log2FC_Rep2), log2FC_Rep2 > 0) %>% pull(pair_id) %>% unique()
rep2_down <- df_ids %>% filter(!is.na(log2FC_Rep2), log2FC_Rep2 < 0) %>% pull(pair_id) %>% unique()


# make venn diagrams
pairwise_venn_grob <- function(A, B, title, fill = c("firebrick","firebrick1")) {
  A <- unique(A); B <- unique(B)
  grid::grid.grabExpr({
    grid::grid.newpage()
    grid::grid.draw(
      draw.pairwise.venn(
        area1 = length(A),
        area2 = length(B),
        cross.area = length(intersect(A, B)),
        category = c("Replicate 1", "Replicate 2"),
        scaled = TRUE,
        fill = fill, alpha = 0.5, lwd = 2,
        cex = 1.2, cat.cex = 1.0, main = title
      )
    )
  })
}
venn_up   <- pairwise_venn_grob(rep1_up,   rep2_up,   "DE up (log2FC > 0)",
                                fill = c("firebrick","firebrick1"))
venn_down <- pairwise_venn_grob(rep1_down, rep2_down, "DE down (log2FC < 0)",
                                fill = c("steelblue","skyblue"))
grid.newpage()
grid.arrange(venn_up, venn_down, ncol = 2)
#default
(368 + 445) / (368 + 445 + 216 + 106)
#2PCs
(200 + 284) / (200 + 284 + 144 + 83)
#3PCs
(195 + 236) / (195 + 236 + 127 + 64)

```


# Check avg on-target using sceptre DE table

- Because of stricter QC for cells in sceptre, on-target repression and the number of genes which have significant on-target gene repression is considerably lower.

```{r}
# Load "final" sceptre DE tables
Rep1_5X_sceptre_de <- read_csv("/users/lvelten/jbowness/CRISPRi500/HSC/clean_analysis/manuscript/int_files/Rep1_5X_sceptre_de_ext.csv")
Rep1_Rem_sceptre_de <- read_csv("/users/lvelten/jbowness/CRISPRi500/HSC/clean_analysis/manuscript/int_files/Rep1_Rem_sceptre_de_ext.csv")

# subset to on-target
Rep1_5X_sceptre_de_ontarget <- subset(Rep1_5X_sceptre_de, target_gene == response_id)
Rep1_Rem_sceptre_de_ontarget <- subset(Rep1_Rem_sceptre_de, target_gene == response_id)
Rep1_sceptre_de_ontarget <- rbind(Rep1_5X_sceptre_de_ontarget,Rep1_Rem_sceptre_de_ontarget)
summary(Rep1_sceptre_de_ontarget$log_2_fold_change)


```


# Session Info

```{r}
sessionInfo()
```
